---
title: "gapminder-wrangle"
author: "Qiyao Jiang"
date: "2025-06-28"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Ways to Analyze Long format data

## Step1: Prepare your package

We introduce package "tidyverse", a strong package for data processing.

```{r load_library}
library(tidyverse)
```

## Step2: Prepare your data

We read the data using readr::read_csv("readr::" can be omitted if we have written "library(tidyverse)"), and use str(), names(), head(), tail() to have a first glimpse of our data. Here we use gapminder.csv for an example.

```{r block_1}
gapminder <- readr::read_csv("data/gapminder.csv")
str(gapminder)
names(gapminder)
head(gapminder, 5)
```

## Step3: Data processsing

### 1.filter(data=data, your_logical_expression(can have multiple expressions here), ...)

This will select the rows that satisfy your conditions.

```{r block_2}
filter(gapminder, lifeExp<29)
filter(gapminder, country %in% c("China", "Japan"))
```

### 2.group_by(), mutate(), summarize(), arrange() and pipe "\|\>"

For more information, please refer to slides "wrangling.pdf", "grouping.pdf" & "Chapter4.pdf".

```{r group_func}
gapminder |>
  filter(year==2002) |>
  group_by(continent) |>
  mutate(count_pop = sum(pop))

gapminder |>
  group_by(continent, year) |>
  summarize(count_pop = sum(pop)) |>
  ungroup()

gapminder |>
  group_by(continent, year) |>
  summarize(gdp = sum(gdpPercap*pop)) |>
  arrange(gdp, .by_group=FALSE)
```

This is another example of the workflow. Notice that you have to read .tsv files with readr::read.**tsv**, NOT readr::read.**csv**!

```{r an_example}
judgements <- readr::read_tsv("https://biostat2.uni.lu/practicals/data/judgments.tsv", show_col_types=FALSE)
glimpse(judgements)
names(judgements)
```

```{r}
select(judgements, gender, age, condition) |>
  arrange(age)


new_judgements <- judgements |>
  mutate(
    mood_change = mood_post - mood_pre,
    mood_change_norm = abs(mood_change / mean(mood_change, na.rm=TRUE))) |>
  relocate(starts_with("mood")) |>
  arrange(desc(mood_change_norm))
head(new_judgements)
mean(new_judgements$mood_change, na.rm=TRUE)
```

```{r}
co2 = read_csv("data/co2.csv")
head(co2, 10)
names(co2)

gap_2007 <- gapminder |>
  filter(year==2007)
dim(gap_2007)

rj <- right_join(gap_2007, co2, by="country")
rj
```

# Before Data Processing: CLEAN Your Data

This is a terrible wide format data:

```{r}
gap_wide <- read_csv("data/gapminder_wide.csv")
head(gap_wide)
```

We can use pivot_longer()(and pivot_wider() if we like) to transfer it into long format nice and neatly:

```{r}
gap_long <- gap_wide |>
  pivot_longer(
    cols = -c("continent", "country"),
    names_to = c("Variable","Year"),
    names_sep = "_",
    values_to = "Value"
  )  |>
  pivot_wider(
    names_from = Variable,
    values_from = Value
  ) |>
  mutate(
    Year = as.numeric(Year),
    GDP = pop*gdpPercap
  )
head(gap_long)
tail(gap_long)
str(gap_long)
```
Then, we can plot our data! Remember what we have done in the last session about ggplot2?
```{r plot_your_data}

China_verses_Japan_and_America_df <- gap_long |>
  filter(country %in% c("China", "Japan", "United States")) 
China_verses_Japan_and_America_df

ggplot(China_verses_Japan_and_America_df) +
  geom_point(aes(x=Year, y=gdpPercap, color=country), alpha=0.7, size=2.5) +
  geom_line(aes(x=Year, y=gdpPercap, color=country)) +
  labs(x="Year", y="gdpPercap($)", title="GDP Per Cap of China, Japan and US") +
  theme_bw() +
  theme(legend.title = element_blank())

ggplot(China_verses_Japan_and_America_df) +
  geom_point(aes(x=Year, y=GDP, color=country), alpha=0.7, size=2.5) +
  geom_line(aes(x=Year, y=GDP, color=country)) +
  labs(x="Year", y="GDP($)", title="GDP of China, Japan and US") +
  theme_bw() +
  theme(legend.title = element_blank())

```